Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-14 paper2.qmd
section:theoretical; process:write;figures,code

$$
RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
$$ {#eq-mat3}

This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of ${n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. The matrix's dimension results from the conceptual-population model's assumption that researchers have access to data capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population. The coefficient is defined as follows:

@@

::: {#fig-cj09_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} \\
\end{aligned}
$$

@@

@fig-cj11 illustrates the sample-comparison model, incorporating the sample mechanism. Following the convention of @McElreath_2020 [pp. 499-516], the [DAG -@fig-cj11_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population. The DAG also depicts the conceptual-population outcome $O^{cp}_{RPUV}$ is now an unobserved variable, emphasizing the researchers' inability to access such an outcome.

::: {#fig-cj11 layout-nrow=2}

::: {#fig-cj11_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} \\
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=84%}

@@

::: {#fig-cj13 layout-nrow=2}

::: {#fig-cj13_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} \\
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=100%}

@@

>>>>

$$
RPUV_{n \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
$$ {#eq-mat3}

This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. The matrix's dimension results from the conceptual-population model's assumption that researchers have access to data capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population. The coefficient is defined as follows:

@@

::: {#fig-cj09_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} 
\end{aligned}
$$

@@

@fig-cj12 illustrates the sample-comparison model, incorporating the sample mechanism. Following the convention of @McElreath_2020 [pp. 499-516], the [DAG -@fig-cj12_dag] shows several conditioned variables $(S_{R}$, $S_{P}$, $S_{U}$, and $S_{V})$ that causally influence the sample-comparison outcome $O^{sc}_{RPUV}$, reflecting how the outcome is a sample drawn from the conceptual population. The DAG also depicts the conceptual-population outcome $O^{cp}_{RPUV}$ is now an unobserved variable, emphasizing the researchers' inability to access such an outcome.

Concatenating the matrices

::: {#fig-cj12 layout-nrow=2}

::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=84%}

@@

::: {#fig-cj14 layout-nrow=2}

::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J} 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=100%}

@@

::: {#fig-cj15 layout-ncol=2}

::: {#fig-cj15_stat fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S_{RPUV} \\
  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
  e_{I} & \sim \text{Normal}(0,1) \\
  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\
  e_{J} & \sim \text{Normal}(0,1)
\end{aligned}
$$

Statistical model
:::

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \dsep \{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \dsep \{ e_{JK}, e_{J} \} \\
  e_{JK} & \dsep e_{J}
\end{aligned}
$$
